Iterative Monomorphisation

### 2024-01-30 iterative_monomorphisation.tex
switched styles and improved tables

\documentclass{article}

\usepackage{soul}
\usepackage{caption}
\usepackage{booktabs}
\usepackage{array}
\usepackage{multirow}
\usepackage{algorithm}
\usepackage{algpseudocode}
\usepackage{amsmath}
\usepackage{authblk}

@@

\title{Iterative Monomorphisation}
\author{Jasmin Blanchette, Sascha B\"ohme, Tanguy Bozec}
\affil{[add affiliation]}
\date{[date]}
\maketitle
\begin{abstract}
   [Write abstract]
\end{abstract}

% target page number: 2-2.5
\section{Introduction}
%context and motivation (automatic theorem proving, better performance)\\
Automatic theorem provers have many applications, and are particularly useful as backends to proof assistants. Speed is one of the main limitations of the use of theorem provers in this setting. Indeed, a prover can often be expected to run for no more than a few seconds for a given instance of a problem.

%what we are trying to do (monomorphisation)\\
This paper presents an iterative method, first introduced in [sacha's thesis], that allows for higher order polymorphic problems to be approximated monomorphically. This algorithm aims to acheive the following:
   \begin{itemize}
      \item allow for existing higher order monomorphic theorem provers to be extended to higher order polymorphic languages
      \item improve the performance of existing higher order polymorphic theorem provers on polymorphic problems
   \end{itemize}

It is therefore aimed at being practically useful, to this end it forgoes completeness, this is the main theoretical weakness of this algorithm.

%basic idea of the algorithm

A given problem is seen as a set of polymorphic clauses. Type variables of the problem are quantified at the clause level. The type arguments of the problem are collected and the uninstantiated type arguments are matched against those that contain no type variables. This yields new type arguments, both uninstantitated and ground, the process is then repeated making use of the newly derived type arguments.

The new type arguments that have been derived are used to create new clauses, without type variables.
    A final step is required as the newly generated problem, whilst it does not contain type variables, still makes use of type constructors. Those can be transparently erased through the mangling method presented in [encoding monomorphic and polymorphic types].

%overview of the paper

This paper will first give some general background and describe the logical framework. Then a high-level mathematical description of the algorithm will be presented. A more detailed low-level description of the algorithm corresponding to its actual implementation will follow.
       Finally, an evaluation of the algorithm's implementation in zippeposition will be discussed.

@@

Definitions :
\begin{itemize}
   \item type arguments % part of the logic
   \item ground and uninstantiated types %monomorphic and polymorphic
   \item matching % being overly precise probably uncecessary
   \item term, literal, clause  % also part logic
\end{itemize}

@@

\begin{algorithm}[tbh]
\begin{algorithmic}[1]
\Function{Type arguments}{Literal: lit}

@@

\begin{table}[H]
\caption{Monomorphic type argument generation}
\centering\begin{tabular}{@{}llc*{9}{>{\centering\arraybackslash}p{1.5em}}@{}}
   && \multicolumn{3}{c}{floor 1} & \multicolumn{3}{c}{floor 3} & \multicolumn{3}{c}{floor 9}\\
   \cmidrule(lr){3-5} \cmidrule(lr){6-8} \cmidrule(lr){9-11}
   && \multicolumn{3}{c}{multiplier} & \multicolumn{3}{c}{multiplier} & \multicolumn{3}{c}{multiplier}\\
   && 0.0 & 1.0 & 2.0 & 0.0 & 1.0 & 2.0 & 0.0 & 1.0 & 2.0\\
   \cmidrule(lr){3-5} \cmidrule(lr){6-8} \cmidrule(lr){9-11}
   \multirow{4}{1.5em}{cap}
   & \multicolumn{1}{l|}{25}   & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 \\
   & \multicolumn{1}{l|}{50}   & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 \\
   & \multicolumn{1}{l|}{100}  & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 \\
   & \multicolumn{1}{l|}{none} & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 \\
\end{tabular}
\end{table}

@@

\begin{table}[H]
\caption{Polymorphic type argument generation}
\centering\begin{tabular}{@{}llc*{9}{>{\centering\arraybackslash}p{1.5em}}@{}}
   && \multicolumn{3}{c}{floor 0} & \multicolumn{3}{c}{floor 3} & \multicolumn{3}{c}{floor 6}\\
   \cmidrule(lr){3-5} \cmidrule(lr){6-8} \cmidrule(lr){9-11}
   && \multicolumn{3}{c}{multiplier} & \multicolumn{3}{c}{multiplier} & \multicolumn{3}{c}{multiplier}\\
   && 0.0 & 0.5 & 1.0 & 0.0 & 0.5 & 1.0 & 0.0 & 0.5 & 1.0\\
   \cmidrule(lr){3-5} \cmidrule(lr){6-8} \cmidrule(lr){9-11}
   \multirow{4}{1.5em}{cap}
   & \multicolumn{1}{l|}{10} 
   & \multirow{4}{1em}{0}       & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 \\
   & \multicolumn{1}{l|}{20}   && 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 \\
   & \multicolumn{1}{l|}{50}   && 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 \\
   & \multicolumn{1}{l|}{none} && 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 \\
\end{tabular}
\end{table}

@@

\begin{table}[H]
\caption{Clause generation}
\centering\begin{tabular}{@{}llc*{9}{>{\centering\arraybackslash}p{1.5em}}@{}}
   && \multicolumn{3}{c}{500 cap} & \multicolumn{3}{c}{2000 cap} & \multicolumn{3}{c}{no cap}\\
   \cmidrule(lr){3-5} \cmidrule(lr){6-8} \cmidrule(lr){9-11}
   && \multicolumn{3}{c}{multiplier} & \multicolumn{3}{c}{multiplier} & \multicolumn{3}{c}{multiplier}\\
   && 1.0 & 2.0 & 3.0 & 1.0 & 2.0 & 3.0 & 1.0 & 2.0 & 3.0\\
   \cmidrule(lr){3-5} \cmidrule(lr){6-8} \cmidrule(lr){9-11}
   \multirow{4}{4.2em}{e-timeout}
   & \multicolumn{1}{l|}{1 s}  & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 \\
   & \multicolumn{1}{l|}{5 s}  & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 \\
   & \multicolumn{1}{l|}{10 s} & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 \\
   & \multicolumn{1}{l|}{none} & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 \\
\end{tabular}
\end{table}

The next table gathers the bounds linked to substitutions generation, the first one limits the number of monomorphising substitutions that are generated per clause during the final phase. The second one gives an absolute cap on the overall number of substitutions generated at each clause during the type argument generation.

\begin{table}[H]
\caption{Substitution generation}
\centering\begin{tabular}{@{}lc*{4}{>{\centering\arraybackslash}p{1.5em}}@{}}
   \toprule
   && \multicolumn{4}{c}{mono subst}\\
   && 2 & 5 & 7 & 10\\
   \cmidrule(lr){3-6}
   \multirow{4}{4.1em}{subst cap}
   & \multicolumn{1}{l|}{50}   & 0 & 0 & 0 & 0\\
   & \multicolumn{1}{l|}{100}  & 0 & 0 & 0 & 0\\
   & \multicolumn{1}{l|}{500}  & 0 & 0 & 0 & 0\\
   & \multicolumn{1}{l|}{none} & 0 & 0 & 0 & 0\\

@@

\begin{table}[H]
% not super happy with caption
\caption{Problem depth}
\centering\begin{tabular}{@{}lc*{4}{>{\centering\arraybackslash}p{1.5em}}@{}}
   \toprule
   && \multicolumn{4}{c}{E call step}\\
   && 0 & 15 & 45 & 90\\
   \cmidrule(lr){3-6}
   \multirow{4}{3.3em}{loop nb}
   & \multicolumn{1}{l|}{2} & 0 & 0 & 0 & 0\\
   & \multicolumn{1}{l|}{3} & 0 & 0 & 0 & 0\\
   & \multicolumn{1}{l|}{4} & 0 & 0 & 0 & 0\\
   & \multicolumn{1}{l|}{5} & 0 & 0 & 0 & 0\\

@@

% final tests to make sure everything plays well together
\begin{table}[H]
\caption{Problem size refinement}
\centering\begin{tabular}{@{}llcccc@{}}
   \toprule
   && \multicolumn{4}{c}{substitution generation} \\
   && \multicolumn{2}{c}{old} & \multicolumn{2}{c}{new} \\
   \cmidrule(lr){3-4} \cmidrule(lr){5-6}
   && \multicolumn{2}{c}{problem depth} &\multicolumn{2}{c}{problem depth}\\
   && old & new & old & new \\
   \cmidrule(lr){3-4} \cmidrule(lr){5-6}
   \multirow{2}{7.5em}{clause generation}
   &\multicolumn{1}{l|}{old} & 0  & 0 & 0 & 0 \\
   &\multicolumn{1}{l|}{new} & 0  & 0 & 0 & 0\\
   \bottomrule
\end{tabular}
\end{table}

\begin{table}[H]
\caption{Type argument generation refinement}
\centering\begin{tabular}{@{}llcc@{}}
   \toprule
   && \multicolumn{2}{c}{mono bounds} \\
   && old & new\\
   \cmidrule(r){3-4}
   \multirow{2}{5.4em}{poly bounds}
   &\multicolumn{1}{l|}{old} & 0  & 0 \\
   &\multicolumn{1}{l|}{new} & 0  & 0 \\
   \bottomrule
\end{tabular}
\end{table}

\begin{table}[H]
\caption{Final bounds refinement}
\centering\begin{tabular}{@{}llcc@{}}
   \toprule
   && \multicolumn{2}{c}{type argument bounds} \\
   && old & new\\
   \cmidrule(r){3-4}
   \multirow{2}{5.4em}{problem size bounds}
   &\multicolumn{1}{l|}{old} & 0  & 0 \\
   &\multicolumn{1}{l|}{new} & 0  & 0 \\
   \bottomrule
\end{tabular}
\end{table}

>>>>

\documentclass[]{ceurart}

\sloppy

\usepackage{soul}
\usepackage{caption}
\usepackage{booktabs}
\usepackage{listings}
\usepackage{array}
\usepackage{multirow}
\usepackage{algorithm}
\usepackage{algpseudocode}
\usepackage{amsmath}

\lstset{breaklines=true}

@@

\copyrightyear{2024}
\copyrightclause{standard copyright text (cc logo is placeholder)}

\conference{not a conference}

\title{Iterative Monomorphisation}

\author[1]{Jasmin Blanchette}[%
email=jasmin.blanchette@ifi.lmu.de,
url=https://www.tcs.ifi.lmu.de/mitarbeiter/jasmin-blanchette_de.html,
]
\address[1]{lmu}
\author{Sascha B\"ohme}[%
email=sascha.boehme@posteo.de,
]
\author[1,2]{Tanguy Bozec}[%
email=tanguy.bozec@ens-paris-saclay.fr,
]
\address[2]{ens paris-saclay}

\begin{abstract}
   Monomorphisation methods allows the extension of some higher order theorem provers to polymorphic logics. This paper presents an iterative approach to monomorphisation with two main goals: creating an efficient method for extending higher order provers to polymorphic logics and improving the performance of existing provers which natively support polymorphism.

   We first present the iterative monomorphisation algorithm, before discussing its implementation in Zipperposition. Finally, the implementation is evaluated as an alternative to native polymorphism and as a means to strenghen Zipperposition.
\end{abstract}

\begin{keywords}
   monomorphisation\sep
   higher order theorem proving \sep
   automatic theorem proving
\end{keywords}

\maketitle

% target page number: 2-2.5
\section{Introduction}

This paper will first give some general background and describe the logical framework. Then a high-level description of the algorithm will be presented. A more detailed low-level description of the algorithm corresponding to its implementation will follow.

Finally, the evaluation of the implementation in Zipperposition will be discussed.

\subsection{Context and motivation}
Automatic theorem provers have many applications, and are particularly useful as backends to proof assistants. Speed is one of the main limitations of the use of theorem provers in this setting. Indeed, a prover can often be expected to run for no more than a few seconds for a given instance of a problem.

%what we are trying to do (monomorphisation)\\
This paper presents an iterative method, first introduced in [sacha's thesis], that allows for higher order polymorphic problems to be approximated monomorphically. This is done through the generation of increasingly large monomorphic type arguments which are used to instantiate the type variables of the original problem. This algorithm aims to acheive the following:
   \begin{itemize}
      \item allow for existing higher order monomorphic theorem provers to be extended to higher order polymorphic languages
      \item improve the performance of existing native polymorphic theorem provers on polymorphic problems
   \end{itemize}

It is therefore aimed at being practically useful, to this end it forgoes completeness, this is the main theoretical weakness of this approach.

%basic idea of the algorithm
\subsection{Algorithm overview}
A given problem is seen as a set of polymorphic clauses. Type variables of the problem are quantified at the clause level. The type arguments of the problem are collected and the polymorphic type arguments are matched against the monomorphic ones. This yields new type arguments, both polymorphic and monomorphic, the process is then repeated making use of the newly derived type arguments.

After a set number of iterations, the newly derived type arguments are used to create new monomorphic clauses, by instantiating their type variables.

A final step is required as the newly generated problem, whilst it does not contain type variables, still makes use of type constructors. Those can be erased through the mangling method presented in [encoding monomorphic and polymorphic types].

\subsection{Evaluation results}

@@

Definitions :
\begin{itemize}
   \item type arguments % part of the logic
   \item monomorphic and polymorphic types
   \item matching % being overly precise probably uncecessary
   \item term, literal, clause  % also part logic
\end{itemize}

@@

\begin{algorithm}[th]
\begin{algorithmic}[1]
\Function{Type arguments}{Literal: lit}

@@

\begin{table}[th]
\caption{Monomorphic type argument generation}
\centering\begin{tabular}{@{}l*{9}{>{\centering\arraybackslash}p{1.1em}}@{}}
   \toprule
   & \multicolumn{3}{c}{floor 1} & \multicolumn{3}{c}{floor 3} & \multicolumn{3}{c}{floor 9}\\
   \cmidrule(l){2-10}
   \multirow{2}{2em}{cap} & \multicolumn{9}{c}{multiplier} \\
      & 0.0 & 1.0 & 2.0 & 0.0 & 1.0 & 2.0 & 0.0 & 1.0 & 2.0\\
    \cmidrule(lr){2-4} \cmidrule(lr){5-7} \cmidrule(l){8-10}
    25   & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 \\
    50   & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 \\
    100  & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 \\
    none & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 \\
    \bottomrule
\end{tabular}
\end{table}

@@

\begin{table}[th]
\caption{Polymorphic type argument generation}
\centering\begin{tabular}{@{}l*{9}{>{\centering\arraybackslash}p{1.1em}}@{}}
   \toprule
   & \multicolumn{3}{c}{floor 0} & \multicolumn{3}{c}{floor 3} & \multicolumn{3}{c}{floor 6}\\
   \cmidrule(l){2-10}
   \multirow{2}{2em}{cap} & \multicolumn{9}{c}{multiplier} \\
      & 0.0 & 0.5 & 1.0 & 0.0 & 0.5 & 1.0 & 0.0 & 0.5 & 1.0 \\
    \cmidrule(lr){2-4} \cmidrule(lr){5-7} \cmidrule(l){8-10}
    10   & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 \\
    20   & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 \\
    50   & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 \\
    none & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 \\
    \bottomrule
\end{tabular}
\end{table}

@@

\begin{table}[th]
\caption{Clause generation}
\centering\begin{tabular}{@{}l*{9}{>{\centering\arraybackslash}p{1.1em}}@{}}
   \toprule
   & \multicolumn{3}{c}{cap 500} & \multicolumn{3}{c}{cap 2000} & \multicolumn{3}{c}{no cap}\\
   \cmidrule(l){2-10}
   \multirow{2}{4em}{E timeout} & \multicolumn{9}{c}{multiplier} \\
      & 1.0 & 2.0 & 3.0 & 1.0 & 2.0 & 3.0 & 1.0 & 2.0 & 3.0 \\
    \cmidrule(lr){2-4} \cmidrule(lr){5-7} \cmidrule(l){8-10}
    2s   & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 \\
    5s   & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 \\
    10s  & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 \\
    none & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 \\
    \bottomrule
\end{tabular}
\end{table}
The next table gathers the bounds linked to substitutions generation, the first one limits the number of monomorphising substitutions that are generated per clause during the final phase. The second one gives an absolute cap on the overall number of substitutions generated at each clause during the type argument generation.

\begin{table}[th]
\caption{Substitution generation}
\centering\begin{tabular}{@{}l*{4}{>{\centering\arraybackslash}p{1em}}@{}}
   \toprule
   \multirow{2}{4em}{subst cap} & \multicolumn{4}{c}{mono subst}\\
   & 2 & 5 & 7 & 10\\
   \midrule
   50   & 0 & 0 & 0 & 0\\
   100  & 0 & 0 & 0 & 0\\
   500  & 0 & 0 & 0 & 0\\
   none & 0 & 0 & 0 & 0\\

@@

\begin{table}[th]
% not super happy with caption
\caption{Problem depth}
\centering\begin{tabular}{@{}l*{4}{>{\centering\arraybackslash}p{1em}}@{}}
   \toprule
   \multirow{2}{3em}{loop nb} & \multicolumn{4}{c}{E call step}\\
   & 0 & 15 & 45 & 90\\
   \midrule
   2 & 0 & 0 & 0 & 0\\
   3 & 0 & 0 & 0 & 0\\
   4 & 0 & 0 & 0 & 0\\
   5 & 0 & 0 & 0 & 0\\

@@

% final tests to make sure everything plays well together
\begin{table}[ht]
\caption{Problem size refinement}
\centering\begin{tabular}{@{}lcccc@{}}
   \toprule
   & \multicolumn{4}{c}{substitution generation} \\
   & \multicolumn{2}{c}{old} & \multicolumn{2}{c}{new} \\
   \cmidrule(lr){2-5} 
   \multirow{2}{4.5em}{clause gen} & \multicolumn{4}{c}{problem depth} \\
   & old & new & old & new \\
   \cmidrule(lr){2-3} \cmidrule(lr){4-5}
   old & 0  & 0 & 0 & 0 \\
   new & 0  & 0 & 0 & 0\\
   \bottomrule
\end{tabular}
\end{table}

\begin{table}[ht]
\caption{Type argument generation refinement}
\centering\begin{tabular}{@{}llcc@{}}
   \toprule
   && \multicolumn{2}{c}{mono bounds} \\
   && old & new\\
   \cmidrule(r){3-4}
   \multirow{2}{5.4em}{poly bounds}
   &\multicolumn{1}{l}{old} & 0  & 0 \\
   &\multicolumn{1}{l}{new} & 0  & 0 \\
   \bottomrule
\end{tabular}
\end{table}

\begin{table}[ht]
\caption{Final bounds refinement}
\centering\begin{tabular}{@{}llcc@{}}
   \toprule
   && \multicolumn{2}{c}{type argument bounds} \\
   && old & new\\
   \cmidrule(r){3-4}
   \multirow{2}{5.4em}{problem size bounds}
   &\multicolumn{1}{l}{old} & 0  & 0 \\
   &\multicolumn{1}{l}{new} & 0  & 0 \\
   \bottomrule
\end{tabular}
\end{table}
